---
title: "Differential Interaction Analysis"
author: "Christopher Adelmann"
format: html
editor: visual
df-print: paged

knitr:
  opts_chunk:
    rows.print: 20

project:
  execute-dir: project
  
params:
  out_dir: "."
  analysis_method: "DESeq2" # either edgeR or DESeq2
  metadata_file: "metadata.csv"
  counts_file: "/Volumes/chris_ssd/projects/external/sRNA_mRNA_interactome_hyderabad/sequencing_run_2/light_to_dark/05_postprocess/complete_super_interaction_transcript_counts.gct"
  interactions_file: "/Volumes/chris_ssd/projects/external/sRNA_mRNA_interactome_hyderabad/sequencing_run_2/light_to_dark/05_postprocess/complete_super_interaction_regions.bedpe"
  annotations_file: "/Users/christopherphd/Downloads/Untitled/Syn_final_modified.gff"
  min_inter_contr: 10
  min_n_samples_contr: 2
  min_n_samples_greater_zero: 3
  feature_ids: "ncr0700,psrR1,csiR1,gene856,gene857,gene2199,gene2962"
  padj_thresh: 0.1
---

```{r}
#| include: false

source_code_files = list.files("code", full.names = TRUE)
sapply(source_code_files, source)
```

# Input Parameters

```{r}
#| echo: false
print(str_glue("min_inter_contr: {params$out_dir}"))
print(str_glue("min_inter_contr: {params$min_inter_contr}"))
print(str_glue("min_n_samples_contr: {params$min_n_samples_contr}"))
print(str_glue("min_n_samples_greater_zero: {params$min_n_samples_greater_zero}"))
print(str_glue("feature_ids: {params$feature_ids}"))
print(str_glue("padj_thresh: {params$padj_thresh}"))
```

# Input Parsing

## Parse Metadata, Counts and Setup Analysis

Input metadata. This determines which subset of samples and conditions will be compared.

Input counts. This document contains all count data for the analysis and is filtered by the metadata.

```{r}
#| echo: false

analysis = make_analysis(params)
```

# Differential Interaction Analysis

## Filtering Counts

```{r}
#| echo: false

analysis = get_filtered_dataset(analysis)
```

## Running Analysis

```{r}
#| echo: false

analysis = run_analysis(analysis)
```

## Quality Control

### Input Data and Count Normalization

```{r}
#| echo: false

plot_read_counts(analysis)

plot_sparsity(analysis)

plot_disp_ests(analysis)
```

### Density of Ratios between Condition

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6


degCheckFactors(analysis$norm_counts)
```

### QC Overview and Standard deviation vs. min median of genes

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6

plot_deg_qc(analysis)
plot_ma(analysis)
```

## Results

### Top 20 Counts

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7

plot_deg(analysis)
```

### LFC Shrunken Results

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
summary(analysis)

plot_volcano(analysis)
```

### Sorted Significant Interaction Regions

```{r}
#| include: false
annotation_gff = read_delim(params$annotations_file, delim = "\t", escape_double = FALSE, trim_ws = TRUE, skip = 7, col_names = FALSE)

interactions_bed = read_delim(params$interactions_file, delim = "\t", escape_double = FALSE, trim_ws = TRUE, skip = 2, col_names = FALSE)

interactions_bed = as.data.frame(interactions_bed)
rownames(interactions_bed) = interactions_bed$X7
```

```{r}
#| echo: false
sig_results_df = sig_results(analysis)
sig_results_df

found_sig_res = TRUE

if (nrow(sig_results_df) < 1) {
  print("Did not find significant results. Stopping further analysis.")
  
  found_sig_res = FALSE
}

if (found_sig_res) {
  analysis = set_sig_interactions(analysis, interactions_bed)

  condition_levels = analysis$metadata_df$condition
  
  result_path = str_glue("{params$out_dir}/condition_{condition_levels[1]}_vs_{condition_levels[2]}_padj{params$padj_thresh}_sig_results.tsv")

  write.table(sig_results_df, result_path, na = "NA", append = TRUE, col.names = TRUE, row.names = TRUE, sep = "\t", quote = TRUE)
}
```

```{r}
#| eval: !expr 'found_sig_res'
#| echo: false
#| fig-width: 8
#| fig-height: 8



sample_bound = 30
z_score_counts = select_ordered_norm_counts(analysis, n = sample_bound)
padj = select_ordered_padj(analysis, n = sample_bound)
sample_bound = min(sample_bound, nrow(z_score_counts))

pheatmap(z_score_counts, cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=FALSE, annotation_col=analysis$metadata_df, annotation_row = padj, main = str_glue("Top {sample_bound} Variance Stabilized\nZ-Scored Counts of Sig. Interactions"))

```

### Significant Interactions of Interest within Results

```{r}
#| eval: !expr found_sig_res
#| echo: false
#| results: asis
#| fig-show: asis


feature_id_list = unlist(strsplit(params$feature_ids, ","))

for (feature_id_str in feature_id_list) {
  cat(str_glue("#### Target interaction {feature_id_str}\n\n"))
  
  # 1) subset the BED df, check rows
  target_rows = interactions_bed[
    grepl(feature_id_str, interactions_bed$X12) |
    grepl(feature_id_str, interactions_bed$X13),
  ]
  
  if (nrow(target_rows) < 1) {
    cat("*No interactions found*\n\n")
    next
  }
  
  # 2) pull out the interaction IDs
  target_interactions = target_rows$X7
  
  # 3) subset your normalized‐counts matrix, check rows
  target_norm_counts = analysis$norm_counts[rownames(analysis$norm_counts) %in% target_interactions, , drop = FALSE]
  if (nrow(target_norm_counts) < 1) {
    cat("*No normalized counts found for those interactions*\n\n")
    next
  }
  
  # 4) order by average abundance
  target_norm_counts = target_norm_counts[
    order(rowMeans(target_norm_counts), decreasing = TRUE),
  , drop = FALSE]
  
  # 5) print tables if you need to inspect them
  target_norm_counts
  target_rows[target_rows$X7 %in% rownames(target_norm_counts), ]
  
  p = pheatmap(
    target_norm_counts,
    cluster_rows    = FALSE,
    cluster_cols    = FALSE,
    show_rownames   = TRUE,
    annotation_col  = analysis$metadata_df,
    main            = str_glue("Log counts of sig. interactions of {feature_id_str}")
  )
  print(p)
  
  cat("\n\n")
}

```

### Annotated Results

```{r}
#| echo: false
#| eval: !expr found_sig_res

annotated_sig_interactions = annotate_sig_interactions(analysis, annotation_gff)


annotated_sig_interactions

result_path = str_glue("{params$out_dir}/condition_{condition_levels[1]}_vs_{condition_levels[2]}_padj{params$padj_thresh}_sig_results_annotated.bedpe")

write.table(annotated_sig_interactions, result_path, na = "NA", append = TRUE, col.names = TRUE, row.names = FALSE, sep = "\t", quote = TRUE)

# Export to bed file
```

### Interaction Count by unambiguous Bio-Type

```{r}
#| echo: false
#| eval: !expr found_sig_res

unambiguous_results = prepare_unambiguous_results(annotated_sig_interactions)

# intra- (same-gene) interactions
intra_mat = get_biotype_matrix(unambiguous_results, "intra")
plot_biotype_heatmap(
  intra_mat,
  "Cross-tabulation of Intramolecular Interaction\nBiotype 1 vs Biotype 2"
)

# inter- (different-gene) interactions
inter_mat = get_biotype_matrix(unambiguous_results, "inter")
plot_biotype_heatmap(
  inter_mat,
  "Cross-tabulation of Intermolecular Interaction\nBiotype 1 vs Biotype 2"
)
```
