---
title: "Differential RNA-RNA Interaction Enrichment Analysis"
author: "Christopher Adelmann"
format:
  html:
    embed-resources: true
editor: visual
df-print: paged

knitr:
  opts_chunk:
    rows.print: 20

project:
  execute-dir: project

params:
  out_dir: "results"
  analysis_method: "both"
  metadata_file: "data/metadata.csv"
  counts_file: "data/complete_super_interaction_transcript_counts.gct"
  interactions_file: "data/complete_super_interaction_regions.bedpe"
  annotations_file: "/Users/christopherphd/Downloads/Syn_final_modified.gff"
  rnanue_results_dir: "/Volumes/chris_ssd_2/external/hyderabad_synechocystis/results_dark"
  detect_dir: null
  analyze_dir: null
  postprocess_dir: null
  primary_contrast:
    value:
      case_role: "treatment"
      control_role: "ligation_control"
  qc_roles:
    value:
      - "no_ligation_control"
  normalization:
    value:
      sample_exposure: "split_reads"
      pair_background: true
      offset_pseudocount: 0.5
  count_handling:
    value:
      edgeR: "numeric"
      DESeq2: "round"
  high_confidence:
    value:
      rnanue_padj_max: 0.1
      max_no_ligation_count: 0
      min_pair_background_available_fraction: 1
      excluded_only_coverage_profiles:
        - "broad_diffuse"
  annotation:
    value:
      enabled: true
      gff_file: "/Users/christopherphd/Downloads/Syn_final_modified.gff"
      feature_types:
        - "gene"
        - "sRNA"
        - "tRNA"
        - "rRNA"
        - "ncRNA"
        - "transcript"
        - "TU"
        - "CDS"
        - "mobile_genetic_element"
      ignore_feature_types:
        - "region"
        - "sequence_feature"
      match_attributes:
        - "ID"
        - "Name"
        - "gene"
        - "locus_tag"
        - "Alias"
        - "Parent"
      biotype_attributes:
        - "gene_biotype"
        - "biotype"
        - "gbkey"
        - "type"
      seqname_normalization: "auto_strip_version"
      strand_policy: "both"
      min_overlap_bp: 1
      target_ids:
        - "ncr0700"
        - "psrR1"
        - "csiR1"
        - "gene856"
        - "gene857"
      target_result_sets:
        - "high_confidence"
        - "concordant_same_direction"
        - "edgeR_significant"
        - "DESeq2_significant"
  batch_correction:
    value:
      mode: "auto"
      columns:
        - "batch"
      on_confounded: "fail"
      visualization_remove_batch: true
  min_inter_contr: 10
  min_n_samples_contr: 2
  min_n_samples_greater_zero: 3
  feature_ids: "ncr0700,psrR1,csiR1,gene856,gene857,gene2199,gene2962"
  padj_thresh: 0.1
  benchmark_metadata:
    value: {}
---

```{r}
#| include: false
source("code/load_libraries.R")
source_code_files <- setdiff(
  list.files("code", pattern = "\\.R$", full.names = TRUE),
  c("code/load_libraries.R", "code/check_environment.R")
)
invisible(lapply(source_code_files, source))
```

# Analysis Overview

This report tests whether RNAnue interaction counts are enriched in treatment samples relative to ligation controls. No-ligation samples are QC-only and are excluded from the primary model.

```{r}
#| echo: false
analysis <- make_analysis(params)
analysis <- run_analysis(analysis)
```

Model-fitting and post-processing warnings are written to `analysis_warnings.tsv` under `out_dir`.

Resolved inputs, contrast, model, filtering, and normalization settings:

```{r}
#| echo: false
knitr::kable(data.frame(
  parameter = c(
    "analysis_method",
    "primary contrast",
    "detect_dir",
    "analyze_dir",
    "padj_threshold",
    "design_formula",
    "batch_status",
    "batch_columns_used",
    "count filter",
    "sample exposure",
    "pair background offset"
  ),
  value = c(
    analysis$params$analysis_method,
    paste0(analysis$params$case_role, " vs ", analysis$params$control_role),
    analysis$params$detect_dir,
    analysis$params$analyze_dir,
    analysis$params$padj_threshold,
    paste(deparse(analysis$design_formula), collapse = ""),
    analysis$design_diagnostics$batch_status[1],
    analysis$design_diagnostics$used_batch_columns[1],
    paste0(
      "count >= ", analysis$params$min_inter_contr,
      " in >= ", analysis$params$min_n_samples_inter_contr,
      " samples and >0 in >= ", analysis$params$min_n_samples_greater_zero,
      " samples"
    ),
    analysis$params$sample_exposure,
    analysis$params$pair_background
  )
))
```

# Sample Roles And Model Design

Sample metadata. Treatment and ligation-control samples enter the model; no-ligation samples are QC-only.

```{r}
#| echo: false
knitr::kable(analysis$metadata)
```

Usable batch columns are added as fixed effects. Confounded batch columns stop the render before model fitting.

Design diagnostics: formula, batch handling, matrix rank, and tested coefficient.

```{r}
#| echo: false
knitr::kable(analysis$design_diagnostics)
```

Batch balance by sample, role, condition, and batch value.

```{r}
#| echo: false
if (nrow(analysis$batch_balance) > 0) {
  knitr::kable(analysis$batch_balance)
} else {
  cat("No usable or requested batch columns were present in model metadata.")
}
```

# Sample-Level Quality Control

QC plots show read depth, interaction library size, interaction yield, and split-read fraction by sample role. `mapped_read_count` uses RNAnue `mapped_reads` when available, otherwise `splits + singletons`; `interaction_library_sum` is the per-sample sum of the interaction-count matrix.

```{r}
#| echo: false
sample_qc_plot_list <- sample_qc_plots(analysis)
```

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
print(sample_qc_plot_list$mapped_read_count)
```

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
print(sample_qc_plot_list$interaction_library_size)
```

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
print(sample_qc_plot_list$mapped_vs_interaction_sum)
```

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
print(sample_qc_plot_list$split_fraction)
```

No-ligation QC output: `no_ligation_qc.tsv`.

# Count Matrix And Normalization

Zero-count fraction per sample, ordered by role. High sparsity can reduce power and affect dispersion estimates.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
print(plot_sparsity_overview(analysis))
```

PCA of all filtered model interactions after log2 offset normalization.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
print(plot_pca_assessment(analysis, "all_model_filtered"))
```

Pairwise Spearman correlations on log2 offset-normalized counts, ordered by role.

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 6
print(plot_sample_correlation_heatmap(analysis))
```

Pair background is available when both interaction arms map to transcript-count features in a model sample. A fraction of `1` means all model samples used the pair-background offset for that interaction.

The plots summarize pair-background availability and final log-offset variation.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
plot_offset_diagnostics(analysis)
```

Normalization diagnostics output: `normalization_diagnostics.tsv`.

# Differential Enrichment

Top annotated interactions for each DIA method. p-values and FDR come from the method-specific negative-binomial model; RNAnue and QC fields are included for context.

```{r}
#| echo: false
for (method in names(analysis$method_results)) {
  cat("## ", method, "\n\n", sep = "")
  method_result <- analysis$method_results[[method]]
  result_table <- method_result$annotated_results
  cat(
    "Significant interactions at FDR <= ",
    analysis$params$padj_threshold,
    ": ",
    sum(result_table$significant, na.rm = TRUE),
    "\n\n",
    sep = ""
  )
  print(knitr::kable(head(result_table, 20)))
}
```

MA plots show effect size versus abundance. Volcano plots show effect size versus FDR evidence. Highlighted points are the high-confidence tier.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
for (method in names(analysis$method_results)) {
  print(plot_method_ma(analysis, method))
  print(plot_method_volcano(analysis, method))
}
```

# DIA Method Agreement

Counts of edgeR/DESeq2 agreement classes. Concordant same-direction calls are stronger candidates, but still only enrichment evidence.

```{r}
#| echo: false
if (!is.null(analysis$concordance)) {
  knitr::kable(table(analysis$concordance$concordance_class))
} else {
  cat("Method concordance requires analysis_method: both.")
}
```

edgeR versus DESeq2 log2 fold changes for the same interactions.

```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 6
print(plot_concordance(analysis))
```

Top agreement records with both methods' statistics, result class, and RNAnue sample-level padj entries when available.

```{r}
#| echo: false
if (!is.null(analysis$concordance)) {
  knitr::kable(head(analysis$concordance, 30))
}
```

# High-Confidence Candidates

High-confidence candidates pass all configured gates: edgeR, DESeq2, RNAnue support, no-ligation QC, pair-background availability, and coverage-profile filters.

Candidate outputs include a full TSV, paired BEDPE, and region-level BED.

Evidence ledger for the high-confidence tier.

```{r}
#| echo: false
high_confidence <- analysis$high_confidence_results
cat(
  "High-confidence interactions: ",
  nrow(high_confidence),
  "\n\n",
  sep = ""
)
if (nrow(high_confidence) > 0) {
  knitr::kable(head(high_confidence, 30))
} else {
  cat("No interactions passed all high-confidence gates.")
}
```

# Result Stability

Significant-result PCAs are exploratory QC. They are selection-biased because features come from the same differential test.

PCA diagnostics: computed and skipped feature sets.

```{r}
#| echo: false
knitr::kable(analysis$postprocess$pca_diagnostics)
```

PCA panels for significant, concordant, and high-confidence feature subsets.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
for (pca_set in setdiff(names(analysis$postprocess$pca_scores), "all_model_filtered")) {
  print(plot_pca_assessment(analysis, pca_set))
}
```

# Significant Interaction Heatmaps

Row-scaled log2 offset-normalized counts for significant interaction sets.

Column annotations follow sample role; row annotations show result class and high-confidence status.

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 8
for (heatmap_set in c("edgeR_significant", "DESeq2_significant", "concordant_same_direction", "high_confidence")) {
  draw_significant_complex_heatmap(analysis, heatmap_set, top_n = 80)
}
```

# Normalized Count Profiles

Normalized counts for top called interactions. Points are samples; the black crossbar is the group median.

Source data output: `significant_normalized_counts.tsv`.

Large within-role spread or single-sample dominance should reduce confidence.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
for (count_set in c("edgeR_significant", "DESeq2_significant", "concordant_same_direction", "high_confidence")) {
  print(plot_significant_normalized_counts(analysis, count_set, top_n = 20))
}
```

Candidate stability by interaction: group means, CVs, prevalence, dominance, treatment-control separation, and no-ligation burden.

```{r}
#| echo: false
knitr::kable(head(analysis$postprocess$candidate_stability, 30))
```

RNAnue metric summary by result class.

```{r}
#| echo: false
knitr::kable(analysis$postprocess$rnanue_metric_group_summary)
```

# RNAnue Support Agreement

Primary DIA FDR versus RNAnue treatment-sample support. Rows without finite DIA adjusted p-values are excluded; rows without RNAnue padj remain in the source table.

RNAnue support is summarized from each treatment sample's minimum native padj. Outputs: `dia_rnanue_padj_comparison.tsv` and `dia_rnanue_discordance.tsv`.

The density scatter shows global agreement and highlights non-background result classes.

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 6
print(plot_dia_rnanue_padj_comparison(analysis))
```

Discordance classes separate shared, DIA-only, RNAnue-only, and unsupported evidence.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
print(plot_dia_rnanue_discordance(analysis))
```

# Partner Recurrence

Partner-recurrence output: `partner_recurrence.tsv`. Recurrent partners can reflect hubs, abundance, annotation ambiguity, or repeated local clusters.

Top recurrent partners and their high-confidence counts.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
print(plot_partner_recurrence(analysis))
```

# Genomic Partner Annotation

Significant interaction arms are annotated against the configured GFF by coordinate overlap. Partner class comes from the selected GFF-overlap feature, not from RNAnue feature IDs.

Annotation diagnostics: run status, annotated arms, unresolved arms, and seqname normalization.

```{r}
#| echo: false
knitr::kable(analysis$postprocess$gff_annotation$diagnostics)
```

Arm annotation output: `interaction_arm_gff_annotations.tsv`.

Partner-type matrices count called interactions by unordered partner-class pair.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 6
for (result_set in c("high_confidence", "concordant_same_direction", "edgeR_significant", "DESeq2_significant")) {
  print(plot_partner_type_confusion(analysis, result_set))
}
```

# Target-Gene Views

Configured target IDs are matched against GFF identifiers and RNAnue arm feature IDs.

```{r}
#| echo: false
knitr::kable(head(analysis$postprocess$gff_annotation$target_gene_interactions, 30))
```

Normalized counts for called target interactions.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 8
print(plot_target_gene_normalized_counts(analysis))
```

# Output Manifest

The render writes TSV outputs to `r analysis$params$out_dir` and overwrites files with the same names.

Output manifest for this render.

```{r}
#| echo: false
output_files <- list.files(analysis$params$out_dir, full.names = FALSE)
knitr::kable(data.frame(output_file = output_files))
```
